rbcL ASV clustering

### 2019-11-01 code/rbcl_asv_clustering.Rmd
added mds

```{r libs, include=F}
library(tidyverse)
library(here)
library(DECIPHER)
library(ggforce)
library(Rtsne)
library(ape)
library(ggtree)
library(treeio)
library(dbscan)

@@

@@

tsne <- dists %>% 
  Rtsne(pca=F, perplexity = perplexity)

tsne <- tsne$Y %>% 
  as_tibble()

ggplot(tsne, aes(x=V1,y=V2))+
  geom_point() +
  labs(title="rbcL ASV clustering with t-SNE")
```

```{r genus_tsne, echo=F}
tsne <- tsne %>% 
  add_column(asv_id = colnames(dists)) %>% # label each point by it's ASV ID
  left_join(taxonomy, by="asv_id")

ggplot(tsne, aes(x=V1,y=V2))+
  geom_point(aes(color = genus)) +
  labs(title = "Points labeled by genus")
```

```{r sp_tsne, echo=F}
ggplot(tsne, aes(x=V1,y=V2))+
  geom_point(aes(color = species)) +
  labs(title="Pts labeled by species")
```

# make new categorical to tell ggplot what taxonomy to plot
# include both species and genus level assignments
# lump low abundance Trebouxiophyceae
treb <- c("Myrmecia","Trebouxia","Coccomyxa","Dictyochloropsis","Diplosphaera")
tsne <- tsne %>% 
    mutate(display_taxa = case_when(genus %in% treb ~"other Trebouxiophyceae",
                                  genus == "Stichococcus" ~ "Stichococcus",#########not sure why I need this...
                                  id_level == "species" ~ assignment,
                                  id_level == "genus" ~ assignment))

# order
taxa.order <- c("Chloromonas","Chloromonas_muramotoi","Chloromonas_hohamii","Chloromonas_krienitzii","Chloromonas_Group_B","Raphidonema_nivale","Stichococcus","other Trebouxiophyceae","Chlainomonas", "Chlainomonas_rubra","Sanguina","Sanguina_aurantia","Sanguina_nivaloides" )

ggplot(tsne, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne, is.na(display_taxa)), 
             color = "grey",alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne, !is.na(display_taxa)), 
             aes(color = display_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
  theme_few()
```

tsne.mat <- tsne %>% 
  select(1,2) %>% 
  as.matrix()

dbscan.out <- dbscan::dbscan(tsne.mat, eps=eps) # eps=4 gives best results

tsne <- tsne %>% 
  add_column(otu = as.character(dbscan.out$cluster)) # tsne is in order of asv_id

ggplot(tsne, aes(x=V1, y=V2)) +
  geom_point(aes(color = otu)) +
  labs(title="dbscan clustering")
```

@@

```{r edit-otus, echo=F}
tsne <- tsne %>%
  mutate(otu = case_when((V2 < (-5) & V1 < -17) ~ "9", # chlain "high"
                         (V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "10",# krienitzii
                         TRUE ~ otu))

tsne %>% 
  select(display_taxa, otu) %>% 
  distinct() %>% 
  arrange(otu)

# name each OTU
otu_names <- tibble(otu = as.character(1:10), otu_name = c("Chloromonas spp. C","Sanguina A","Chlainomonas B", "Chloromonas muramotoi E","Chloromonas G","Chloromonas I","Algae H","Trebouxiophyceae J","Chlainomonas F","Chloromonas krienitzii D"))

tsne <- tsne %>% 
  left_join(otu_names, by="otu")

asv_otu_key <- tsne %>% 
  select(asv_id, otu, otu_name)

ggplot(tsne, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne, is.na(display_taxa)), 
             color = "grey",alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne, !is.na(display_taxa)), 
             aes(color = display_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
  theme_few() +
  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name), show.legend = FALSE)

@@

tsne_2 <- tsne %>% 
  left_join(total_rel_abund, by="asv_id")

ggplot(tsne_2, aes(x=V1, y=V2)) +
  ggforce::geom_mark_ellipse(data = tsne, aes(fill=otu_name, label=otu_name), show.legend = FALSE) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_2, is.na(display_taxa)), 
             aes(size = global_rel_abund),
             color = "grey",alpha = 0.7) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_2, !is.na(display_taxa)), 
             aes(color = display_taxa %>% fct_relevel(taxa.order),
                 size = global_rel_abund), 
             alpha = 0.7) +
  labs(title = "rbcL ASV t-SNE clustering", color="ASV taxonomic assignment") +
  theme_few() +
  scale_size(range = c(1, 6), guide="none")
```

@@

tsne_plot +
  theme(legend.position = "none")

# extract the legend
tsne_plot %>% 
  get_legend() %>% 
  as_ggplot

ggsave(here("figs","asv-cluster","asv-legend-dark.pdf"), height = 7, width = 3)
ggsave(here("figs","asv-cluster","asv-cluster-dark.pdf"), height = 5, width = 5)
```

```{r asv_mds, echo=F}

>>>>

```{r libs, include=F}
library(tidyverse)
library(here)
library(DECIPHER)
library(ggforce)
library(Rtsne)
library(ape)
library(ggtree)
library(treeio)
library(dbscan)
library(ggrepel)

@@

set.seed(123)
tsne_full_output <- dists %>% 
  Rtsne(pca=F, perplexity = perplexity)

tsne_pts <- tsne_full_output$Y %>% 
  as_tibble()

ggplot(tsne_pts, aes(x=V1,y=V2))+
  geom_point() +
  labs(title="rbcL ASV clustering with t-SNE")
```

```{r genus_tsne, echo=F}
tsne_pts_taxonomy <- tsne_pts %>% 
  add_column(asv_id = colnames(dists)) %>% # label each point by it's ASV ID
  left_join(taxonomy, by="asv_id")

ggplot(tsne_pts_taxonomy, aes(x=V1,y=V2))+
  geom_point(aes(color = genus)) +
  labs(title = "Points labeled by genus")
```

```{r sp_tsne, echo=F}
ggplot(tsne_pts_taxonomy, aes(x=V1,y=V2))+
  geom_point(aes(color = species)) +
  labs(title="Pts labeled by species")
```

taxa.order <- c("Chloromonas","Chloromonas_muramotoi","Chloromonas_hohamii","Chloromonas_krienitzii","Chloromonas_Group_B","Raphidonema_nivale","Stichococcus","other Trebouxiophyceae","Chlainomonas", "Chlainomonas_rubra","Sanguina","Sanguina_aurantia","Sanguina_nivaloides" )

ggplot(tsne_pts_taxonomy, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_pts_taxonomy, is.na(plot_taxa)), 
             color = "grey", alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_pts_taxonomy, !is.na(plot_taxa)), 
             aes(color = plot_taxa %>% fct_relevel(taxa.order)), 
             alpha = 0.8, size = 3) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
  theme_few()
```

tsne_mat <- tsne_pts_taxonomy %>% 
  select(1,2) %>% 
  as.matrix()

@@

dbscan_out <- dbscan::dbscan(tsne_mat, eps=eps) # eps=4 gives best results

tsne_taxa_otu <- tsne_pts_taxonomy %>% 
  add_column(otu = as.character(dbscan_out$cluster)) # tsne is in order of asv_id

ggplot(tsne_taxa_otu, aes(x=V1, y=V2)) +
  geom_point(aes(color = otu)) +
  labs(title="dbscan clustering")
```

@@

```{r edit-otus, echo=F, fig.width=8, fig.height=6}
tsne_edited_otus <- tsne_taxa_otu %>%
  mutate(otu = case_when((V2 < (-5) & V1 < -17) ~ "9", # chlain "high"
                         (V2 < 14  & V2 > 8 & V1 > (-20) & V1 < (-12)) ~ "10",# krienitzii
                         TRUE ~ otu))

tsne_edited_otus %>% 
  select(plot_taxa, otu) %>% 
  distinct() %>% 
  arrange(otu)

# name each OTU
otu_names <- tibble(otu = as.character(1:10), 
                    otu_name = c("Chloromonas spp. C",
                                 "Sanguina A","Chlainomonas B",
                                 "Chloromonas muramotoi E",
                                 "Chloromonas G",
                                 "Chloromonas I",
                                 "Algae H",
                                 "Trebouxiophyceae J",
                                 "Chlainomonas F",
                                 "Chloromonas krienitzii D"))

tsne_otu_names <- tsne_edited_otus %>% 
  left_join(otu_names, by="otu")

# make a key for downstream use (write to file at end of script)
asv_otu_key <- tsne_otu_names %>% 
  select(asv_id, otu, otu_name)

ggplot(tsne_otu_names, aes(x=V1, y=V2)) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_otu_names, is.na(plot_taxa)), 
             color = "grey",alpha = 0.8, size = 3) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_otu_names, !is.na(plot_taxa)), 
             aes(color = plot_taxa %>% fct_relevel(taxa.order)), alpha = 0.8, size = 3) +
  labs(title = "Points colored by both genus and species", color="ASV taxonomic assignment") +
  theme_few() +
  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), show.legend = FALSE)
```

@@

tsne_rel_abund <- tsne_otu_names %>% 
  left_join(total_rel_abund, by="asv_id")

final_tsne <- ggplot(tsne_rel_abund, aes(x=V1, y=V2)) +
  ggforce::geom_mark_ellipse(aes(fill=otu_name, label=otu_name), 
                             show.legend = FALSE, 
                             alpha=0.2, 
                             linetype="blank", 
                             expand = unit(2, "mm")) +
  # plot points not assigned to genus or species
  geom_point(data = filter(tsne_rel_abund, is.na(plot_taxa)), 
             aes(size = global_rel_abund),
             color = "grey",alpha = 0.7) +
  # color points that were assigned to genus or species
  geom_point(data = filter(tsne_rel_abund, !is.na(plot_taxa)), 
             aes(color = plot_taxa %>% fct_relevel(taxa.order),
                 size = global_rel_abund), 
             alpha = 0.7) +
  labs(title = "rbcL ASV t-SNE clustering", color="ASV taxonomic assignment") +
  ggthemes::theme_few() +
  scale_size(range = c(1, 6), guide="none") ################################################# not changing the scale size
final_tsne
```

@@

# finel_tsne +
#   theme(legend.position = "none")
# 
# # extract the legend
# tsne_plot %>% 
#   get_legend() %>% 
#   as_ggplot 
# 
# ggsave(here("figs","asv-cluster","asv-legend-dark.pdf"), height = 7, width = 3)
# ggsave(here("figs","asv-cluster","asv-cluster-dark.pdf"), height = 5, width = 5)
```

```{r asv_mds, echo=F, fig.width=10, fig.height=8}
mds_coords <- dists %>% 
  cmdscale() %>% # ok to use euclidian dist, no null values when comparing relatedness
  as_tibble() %>% 
  add_column(asv_id = colnames(dists), .before=1)

mds_asv_plot <- mds_coords %>% 
  left_join(taxonomy, by="asv_id") %>% 
  left_join(asv_otu_key, by="asv_id") %>% 
  left_join(total_rel_abund, by="asv_id") %>% 
  filter(!(asv_id %in% c("639", "534", "646"))) # remove two trebouxiophyceae outliers

ggplot(mds_asv_plot, aes(V1, V2)) + 
  ggforce::geom_mark_ellipse(data = mds_asv_plot %>% 
                               filter(otu_name != "Trebouxiophyceae J"), # dont plot the ellipse around other treb
                             aes(fill = otu_name, label=otu_name),
                             show.legend = FALSE, 
                             alpha=0.2, 
                             linetype="blank",
                             expand = unit(2, "mm")) +
  geom_point(aes(alpha = 0.7, color = plot_taxa, size = global_rel_abund)) +
  labs(title = "MDS clustering rbcL ASVs", color="ASV taxonomic assignment") +
  scale_size(range = c(1, 6), guide="none") +
  scale_alpha(guide="none")+
  theme_few() #+
  # geom_text(aes(label = asv_id)) # too see outliers, asv_id=c("639", "534")
```

```{r write_mds, include=F}
